On the entropy of isothermal dusty gas: application to protoplanetary disks

### 2016-08-09 paper.tex
minor intro edits

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas drag introduces a 
wide range of new phenomena that are absent in pure gas
disks. Important examples include the streaming instability
\citep{youdin05a,youdin07b,johansen07} and secular gravitational
instabilities \citep{ward00,youdin11,takahashi14} that lead to
planetesimal formation. More recently, new dusty instabilities have
also been uncovered \citep{loren15,loren16,lamb16}

@@

%particles 
Current state-of-the-art models of dusty protoplanetary
disks directly simulate gas dynamics with explicit
Lagragian dust particles \citep{nelson10,bai10,johansen11,yang14,zhu14,gibbons15,simon16,baruteau16}. In this
approach, the equations of motion for each particle is solved
including dust-gas drag, the strength of which is measured 
by a stopping time $\tstop$ --- the characteristic timescale for the
relative velocity between gas and dust to decay. Thus small $\tstop$
corresponds to tightly-coupled particles.

@@

It is possible to re-formulate the two-fluid equations into that for a
single fluid of for the total density $\rho$ and center-of-mass
velocity $\bm{v}$ of the dusty gas mixture
\citep{laibe14}. The standard hydrodynamic equations for
$\rho$ and $\bm{v}$ are supplemented by evolutionary equations for
the dust-to-gas ratio $\rhod/\rhog$ and the dust-gas velocity
difference $\Delta\bm{v}$. The latter also contributes to
additional source terms in the momentum and energy equations.  
The one-fluid approach is appropriate for 
simulating gas tightly coupled to dust particles with small $\tstop$.  
This can make the two-fluid or Lagragian particles approach
numerically challanging; but is straight forward to model in the
one-fluid framework \citep{price15}.

@@

We consider tightly coupled dust particles with $\tstop\Omega\ll
1$, where $\Omega$ is the disk orbital frequency. We may then use the
`terminal velocity approximation' to set $\Delta \bm{v} = \tstop\nabla
P/\rhog$, where $P$ is the gas pressure \citep{youdin05a,
  jacquet11,laibe14}. We also consider irradiated protoplanetary disks 
in which the temperature is set externally \citep{chiang97,stam08}. We
can then adopt a locally isothermal equation of state $P = c_s^2\rhog$
where $c_s$ is a prescribed sound-speed profile.  This physically
represents short cooling times \citep{lin15}.

@@

This paper is organized as follows. In \S\ref{setup} we list the
equations describing tightly-coupled dusty gas. We show they are in
fact the usual hydrodynamic equations, including a precise 
definition of the entropy isothermal dusty gas. We describe approximate 
equilibrium solutions in \S\ref{eqm}. In \S\ref{limits} we discuss
general stability properties of the dusty problem in various
limits. We use these results to constraint the sharpness of dust edges
in isothermal disks. In \S\ref{linear_problem} we formulate an
explicit stability calcultion to study the dusty VSI, for which we
present numerical results \S\ref{results}. We summarize and discuss
future applications of the dusty/adiabatic gas equivalence in
\S\ref{summary}.

>>>>

\section{Introduction}
%si, sgi as examples
Protoplanetary disks are comprised of a mixture of gas and dust
\citep{chiang10}. Although the overall dust content is small
($\sim 1\%$ by mass), their presence can have profound impacts on the 
gas dynamics and vice versa. Dust-gas drag introduces a 
wide range of new phenomena that are absent in pure gas
disks. Important examples include the streaming instability
\citep{youdin05a,youdin07b,johansen07}, secular gravitational
instabilities \citep{ward00,youdin11,michikoshi12,takahashi14} and
Kelvin-Helmholtz instabilities
\citep{chiang08,barranco09,lee10}. These lead to structure and
planetesimal formation, as well as turbulence. More recently, new
dusty instabilities have also been uncovered
\citep{loren15,loren16,lamb16}

@@

%particles 
Current state-of-the-art models of dusty protoplanetary
disks directly simulate gas dynamics with explicit
Lagragian dust particles \citep{nelson10,bai10,johansen11,yang14,zhu14,gibbons15,simon16,baruteau16}. In this
approach, the equations of motion for each particle is solved
including dust-gas drag, the strength of which is measured 
by a stopping time $\tstop$ --- the decay timescale for the 
relative velocity between gas and dust. Thus small $\tstop$
corresponds to tightly-coupled particles.

@@

It is possible to re-formulate the two-fluid equations into that for a
single fluid of for the total density $\rho$ and center-of-mass
velocity $\bm{v}$ of the dusty gas mixture
\citep{laibe14}. The standard hydrodynamic equations for
$\rho$ and $\bm{v}$ are supplemented by evolutionary equations for
the dust-to-gas ratio $\rhod/\rhog$ and the dust-gas velocity
difference $\Delta\bm{v}$. The latter also contributes to
additional source terms in the momentum and energy equations.  
The one-fluid approach is appropriate for 
simulating gas tightly coupled to dust particles with small $\tstop$.  
This can be numerically challenging for the two-fluid or Lagragian
particles approach; but is straight forward to model in the
one-fluid framework \citep{price15}.

@@

We consider tightly coupled dust particles with $\tstop\Omega\ll
1$, where $\Omega$ is the disk orbital frequency. We may then use the
`terminal velocity approximation' to set $\Delta \bm{v} = \tstop\nabla
P/\rhog$, where $P$ is the gas pressure \citep{youdin05a,
  jacquet11,laibe14}. We also consider irradiated protoplanetary disks 
in which the temperature is set externally \citep{chiang97,stam08}. We
may then adopt a locally isothermal equation of state $P = c_s^2\rhog$
where $c_s$ is a prescribed sound-speed profile.  This physically
represents short cooling times \citep{lin15}.

@@

This paper is organized as follows. In \S\ref{setup} we list the
equations describing tightly-coupled dusty gas. We show they are in 
fact the usual hydrodynamic equations. This leads to a precise 
definition of the entropy of an isothermal dusty gas. We describe approximate 
equilibrium solutions in \S\ref{eqm}. In \S\ref{limits} we discuss
general stability properties of the dusty problem in various
limits. We use these results to constraint the sharpness of dust edges
in isothermal disks. In \S\ref{linear_problem} we formulate an
explicit stability calcultion to study the dusty VSI, for which we
present numerical results in \S\ref{results}. We summarize and discuss
future applications of the dusty/adiabatic gas equivalence in
\S\ref{summary}.
